---
title: "R Notebook for MIP analysis"
author: "LINGYUN XIONG"
date: "13/8/2018"
output:
  html_document:
    df_print: paged
---

Description:

This R script aims to analyze the somatic mutation calling results from MIP targeted sequencing, which serves as a holistic assessment of the quality of tumor DNA sequencing and mutational calling. The results need to be compared with other sequencing methods, and with mutational calling results where paired tumor-normal samples are available. 

## Step One

Read in the input files and pre-process the matrix containing end-point output from sequencing and mutational calling and a list of MSI status information for each tumor sample.

```{r echo=FALSE}
library(ggplot2)
library(reshape2)
df <- as.data.frame(read.csv("~/Documents/Project/Input/input.csv",header=T,sep=","))
mmtx <- as.matrix(df[,61:321])
sample_ID <- as.character(colnames(df)[61:321])
MSI_info <- as.data.frame(read.csv("~/Documents/Project/Input/MSI_status.csv",header=T,sep=","))
colnames(MSI_info) <- c("sample_ID","MSI_status")
df_ID_MSI <- merge(as.data.frame(sample_ID),MSI_info,by="sample_ID",sort=F)
sample_MSI <- as.character(df_ID_MSI$MSI_status)
sample_MSI[which(is.na(sample_MSI))] <- "unknown"
df_ID_MSI$MSI_status <- sample_MSI
```

## Step Two

Annotate each mutation entry for the following classes:

* Indels (coding/non-coding)
* SNVs (coding/non-coding)
    + C>A
    + C>G
    + C>T
    + T>A
    + T>C
    + T>G
    

```{r echo=FALSE}
df_gene_mutation <- df[,c(43,44,46,47,9,1,2,6,4,5,322,323,324)]
mutation_class <- rep("-",3009)
mutation_class[which(nchar(as.character(df_gene_mutation$REF)) > 1 | nchar(as.character(df_gene_mutation$ALT) > 1))] <- "c-Indels"
mutation_class[which(df_gene_mutation$cDNA_position=="-" & (nchar(as.character(df_gene_mutation$REF)) > 1 | nchar(as.character(df_gene_mutation$ALT) > 1)))] <- "nc-Indels"
mutation_class[which(nchar(as.character(df_gene_mutation$REF)) == 1 & nchar(as.character(df_gene_mutation$ALT) == 1))] <- "c-SNVs"
mutation_class[which(df_gene_mutation$cDNA_position=="-" & nchar(as.character(df_gene_mutation$REF)) == 1 & nchar(as.character(df_gene_mutation$ALT) == 1))] <- "nc-SNVs"
df_gene_mutation$mutation_class <- mutation_class
table(mutation_class)
#cDNA_ref <- rep("-",3009)
#cDNA_alt <- rep("-",3009)
#for (i in 1:3009){
#  if(mutation_class[i]=="SNVs"){
#    cDNA_ref[i] <- unlist(strsplit(as.character(df_gene_mutation$Codons)[i],"/"))[1]
#    cDNA_ref[i] <- gsub("[^::A-Z::]","", cDNA_ref[i])
#    cDNA_alt[i] <- unlist(strsplit(as.character(df_gene_mutation$Codons)[i],"/"))[2]
#    cDNA_alt[i] <- gsub("[^::A-Z::]","", cDNA_alt[i])
#  }
#}
#df_gene_mutation$allele_ref <- cDNA_ref
#df_gene_mutation$allele_alt <- cDNA_alt
```

## Step Three

Count the number of entries of various classes in each tumor sample

```{r echo=FALSE}
#####overall distribution#####
###counting total number of mutations in each sample###
total_mutations <- rep(0,length(sample_ID))
for (i in 1:length(sample_ID)){
  total_mutations[i] <- length(which(mmtx[,i]!="."))+length(which(mmtx[,i]=="1/1"))
}
df_ID_MSI$total_mutations <- total_mutations

###counting total number of Indels in each sample###
total_Indels <- rep(0,length(sample_ID))
submtx <- mmtx[which(df_gene_mutation$mutation_class=="c-Indels" | df_gene_mutation$mutation_class=="nc-Indels"),]
for (i in 1:length(sample_ID)){
  total_Indels[i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
df_ID_MSI$total_Indels <- total_Indels

###counting total number of Indels within coding sequence in each sample###
coding_Indels <- rep(0,length(sample_ID))
submtx <- mmtx[which(df_gene_mutation$mutation_class=="c-Indels"),]
for (i in 1:length(sample_ID)){
  coding_Indels[i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
df_ID_MSI$coding_Indels <- coding_Indels

###counting total number of SNVs in each sample###
total_SNVs <- rep(0,length(sample_ID))
submtx <- mmtx[which(df_gene_mutation$mutation_class=="c-SNVs" | df_gene_mutation$mutation_class=="nc-SNVs"),]
for (i in 1:length(sample_ID)){
  total_SNVs[i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
df_ID_MSI$total_SNVs <- total_SNVs

###counting total number of SNVs within coding sequence in each sample###
coding_SNVs <- rep(0,length(sample_ID))
submtx <- mmtx[which(df_gene_mutation$mutation_class=="c-SNVs"),]
for (i in 1:length(sample_ID)){
  coding_SNVs[i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
df_ID_MSI$coding_SNVs <- coding_SNVs

#####gene-level distribution#####
gene_list <- unique(as.character(df$SYMBOL))
gene_total_mutations <- rep(0,length(gene_list))
gene_total_Indels <- rep(0,length(gene_list))
gene_coding_Indels <- rep(0,length(gene_list))
gene_total_SNVs <- rep(0,length(gene_list))
gene_coding_SNVs <- rep(0,length(gene_list))
mtx_gene_sample_Indels <- matrix(0,length(gene_list),length(sample_ID))
mtx_gene_sample_SNVs <- matrix(0,length(gene_list),length(sample_ID))
for (i in 1:length(gene_list)){
  ###counting total number of Indels and SNVs in each gene###
  gene_mutation_class <- as.character(df_gene_mutation[which(df_gene_mutation$SYMBOL==gene_list[i]),]$mutation_class)
  gene_total_mutations[i] <- length(gene_mutation_class)
  gene_total_Indels[i] <- length(gene_mutation_class[which(gene_mutation_class=="c-Indels" | gene_mutation_class=="nc-Indels")])
  gene_coding_Indels[i] <- length(gene_mutation_class[which(gene_mutation_class=="c-Indels")])
  gene_total_SNVs[i] <- length(gene_mutation_class[which(gene_mutation_class=="c-SNVs" | gene_mutation_class=="nc-SNVs")])
  gene_coding_SNVs[i] <- length(gene_mutation_class[which(gene_mutation_class=="c-SNVs")])
  
  ###counting total number of Indels and SNVs of each gene in each sample###
  submtx_gene_Indels <- mmtx[which(df_gene_mutation$SYMBOL==gene_list[i] & (df_gene_mutation$mutation_class=="c-Indels" | df_gene_mutation$mutation_class=="nc-Indels")),]
  submtx_gene_SNVs <- mmtx[which(df_gene_mutation$SYMBOL==gene_list[i] & (df_gene_mutation$mutation_class=="c-SNVs" | df_gene_mutation$mutation_class=="nc-SNVs")),]
  if (!is.null(nrow(submtx_gene_Indels))){
    for (j in seq_along(sample_ID)){
      mtx_gene_sample_Indels[i,j] <- length(which(submtx_gene_Indels[,j]!="."))+length(which(submtx_gene_Indels[,j]=="1/1"))
      }
  }
  if (!is.null(nrow(submtx_gene_SNVs))){
    for (j in seq_along(sample_ID)){
      mtx_gene_sample_SNVs[i,j] <- length(which(submtx_gene_SNVs[,j]!="."))+length(which(submtx_gene_SNVs[,j]=="1/1"))
    }
  }
}
df_gene_summary <- data.frame(gene_list,gene_total_mutations,gene_total_Indels,gene_total_SNVs,gene_coding_Indels,gene_coding_SNVs)
mtx_gene_sample_mutations <- mtx_gene_sample_Indels+mtx_gene_sample_SNVs

#####SNVs: mutation-type-specific distribution#####
mutations_in_each_type <- rep(0,7)
mtx_mtype_sample <- matrix(0,6,length(sample_ID))
###type 1: C>A###
mutations_in_each_type[1] <- length(as.character(df_gene_mutation[which((df_gene_mutation$REF=="C" & df_gene_mutation$ALT=="A") | (df_gene_mutation$REF=="G" & df_gene_mutation$ALT=="T")),]$SYMBOL))
submtx <- mmtx[which((df_gene_mutation$REF=="C" & df_gene_mutation$ALT=="A") | (df_gene_mutation$REF=="G" & df_gene_mutation$ALT=="T")),]
for (i in 1:length(sample_ID)){
  mtx_mtype_sample[1,i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
###type 2: C>G###
mutations_in_each_type[2] <- length(as.character(df_gene_mutation[which((df_gene_mutation$REF=="C" & df_gene_mutation$ALT=="G") | (df_gene_mutation$REF=="G" & df_gene_mutation$ALT=="C")),]$SYMBOL))
submtx <- mmtx[which((df_gene_mutation$REF=="C" & df_gene_mutation$ALT=="G") | (df_gene_mutation$REF=="G" & df_gene_mutation$ALT=="C")),]
for (i in 1:length(sample_ID)){
  mtx_mtype_sample[2,i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
###type 3: C>T###
mutations_in_each_type[3] <- length(as.character(df_gene_mutation[which((df_gene_mutation$REF=="C" & df_gene_mutation$ALT=="T") | (df_gene_mutation$REF=="G" & df_gene_mutation$ALT=="A")),]$SYMBOL))
submtx <- mmtx[which((df_gene_mutation$REF=="C" & df_gene_mutation$ALT=="T") | (df_gene_mutation$REF=="G" & df_gene_mutation$ALT=="A")),]
for (i in 1:length(sample_ID)){
  mtx_mtype_sample[3,i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
###type 4: T>A###
mutations_in_each_type[4] <- length(as.character(df_gene_mutation[which((df_gene_mutation$REF=="T" & df_gene_mutation$ALT=="A") | (df_gene_mutation$REF=="A" & df_gene_mutation$ALT=="T")),]$SYMBOL))
submtx <- mmtx[which((df_gene_mutation$REF=="T" & df_gene_mutation$ALT=="A") | (df_gene_mutation$REF=="A" & df_gene_mutation$ALT=="T")),]
for (i in 1:length(sample_ID)){
  mtx_mtype_sample[4,i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
###type 5: T>C###
mutations_in_each_type[5] <- length(as.character(df_gene_mutation[which((df_gene_mutation$REF=="T" & df_gene_mutation$ALT=="C") | (df_gene_mutation$REF=="A" & df_gene_mutation$ALT=="G")),]$SYMBOL))
submtx <- mmtx[which((df_gene_mutation$REF=="T" & df_gene_mutation$ALT=="C") | (df_gene_mutation$REF=="A" & df_gene_mutation$ALT=="G")),]
for (i in 1:length(sample_ID)){
  mtx_mtype_sample[5,i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
###type 6: T>G###
mutations_in_each_type[6] <- length(as.character(df_gene_mutation[which((df_gene_mutation$REF=="T" & df_gene_mutation$ALT=="G") | (df_gene_mutation$REF=="A" & df_gene_mutation$ALT=="C")),]$SYMBOL))
submtx <- mmtx[which((df_gene_mutation$REF=="T" & df_gene_mutation$ALT=="G") | (df_gene_mutation$REF=="A" & df_gene_mutation$ALT=="C")),]
for (i in 1:length(sample_ID)){
  mtx_mtype_sample[6,i] <- length(which(submtx[,i]!="."))+length(which(submtx[,i]=="1/1"))
}
mutations_in_each_type[7] <- sum(mutations_in_each_type[1:6])
df_mtx_mtype_sample <- data.frame(t(mtx_mtype_sample))
colnames(df_mtx_mtype_sample) <- c("C_to_A","C_to_G","C_to_T","T_to_A","T_to_C","T_to_G")
df_ID_MSI <- cbind(df_ID_MSI,df_mtx_mtype_sample)
```

## Step Four: Results

```{r echo=FALSE}
###plot the total number of Indels and SNVs in each sample###
df <- df_ID_MSI[,c(1,2,3,4,6)]
dfl <- melt(df,id.vars = c("sample_ID","MSI_status","total_mutations"))
p_total_Indels_SNVs <- ggplot(dfl,aes(x=reorder(sample_ID,-total_mutations),y=value,fill=factor(MSI_status),alpha=variable)) +
  geom_bar(stat="identity",size=0.2) + ylab("Frequency of mutations") + xlab("Tumor samples") + 
  ggtitle("Distribution of Indels and SNVs in each tumor sample") +
  scale_fill_manual(values=c("red3","navy","grey50")) +
  scale_alpha_discrete(range=c(0.3,0.8)) + theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

###plot the total number of Indels and SNVs within coding regions in each sample###
df <- df_ID_MSI[,c(1,2,5,7)]
df$total_coding_mutations <- as.vector(df$coding_Indels)+as.vector(df$coding_SNVs)
dfl <- melt(df,id.vars = c("sample_ID","MSI_status","total_coding_mutations"))
p_coding_Indels_SNVs <- ggplot(dfl,aes(x=reorder(sample_ID,-total_coding_mutations),y=value,fill=factor(MSI_status),alpha=variable)) +
  geom_bar(stat="identity",size=0.2) + ylab("Frequency of mutations") + xlab("Tumor samples") + 
  ggtitle("Distribution of Indels and SNVs within coding regions in each tumor sample") +
  scale_fill_manual(values=c("red3","navy","grey50")) +
  scale_alpha_discrete(range=c(0.3,0.8)) + theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

###plot the total number of Indels and SNVs in each gene###
df <- df_gene_summary[,c(1,2,3,4)]
dfl <- melt(df,id.vars = c("gene_list","gene_total_mutations"))
p_gene_Indels_SNVs <- ggplot(dfl,aes(x=reorder(gene_list,-gene_total_mutations),y=value,fill=factor(variable))) +
  geom_bar(stat="identity",size=0.2) + ylab("Frequency of mutations") + xlab("Tumor samples") + 
  ggtitle("Distribution of Indels and SNVs in each gene") +
  scale_fill_manual(values=c("red3","grey")) + theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

###plot C>T mutation against total number of mutations in each sample###
df <- df_ID_MSI[,c(1,2,3,10)]
p_C_T_scatter <- ggplot(df, aes(x=total_mutations,y=C_to_T)) + 
  geom_point() + geom_smooth(method=lm, linetype="dashed", color="red3") +
  ylab("Number of C>T mutations") + xlab("Total number of mutations in each sample") + 
  ggtitle("Scatter plot of C>T transition and total number of mutations in each tumor sample") + theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

###plot total mutations within each gene in each tumor samples###

```

### 1) Histogram of total mutations in each tumor sample (marked with MSI status)
```{r echo=FALSE}
p_total_Indels_SNVs
```

### 2) Histogram of total mutations within coding regions in each tumor sample (marked with MSI status)
```{r echo=FALSE}
p_coding_Indels_SNVs
```

### 3) Histogram of total mutations in each gene
```{r echo=FALSE}
p_gene_Indels_SNVs
```

### 4) Summary of the distribution of total mutations in each mutation type
```{r echo=FALSE}
data.frame(mutation_type=c("C>A","C>G","C>T","T>A","T>C","T>G","total"),total_count_of_mutation=mutations_in_each_type, proportion=round(mutations_in_each_type/mutations_in_each_type[7],2))
```

### 5) Scatter plot of total count of C>T mutations against total mutations in each tumor sample
```{r echo=FALSE}
p_C_T_scatter
```

### 6) Table of the most frequent mutations detected 
```{r echo=FALSE}
df_tmp <- df_gene_mutation[order(df_gene_mutation$presumed.wild.type.count),]
df_tmp[which(df_tmp$presumed.wild.type.count<=250),]
```

### 7) Histograms of total mutations within each gene in each tumor sample
```{r echo=FALSE}
for (i in seq_along(gene_list)){
  df <- data.frame(sample_ID=df_ID_MSI$sample_ID,MSI_status=df_ID_MSI$MSI_status,total_mutations=mtx_gene_sample_mutations[i,],total_Indels=mtx_gene_sample_Indels[i,],total_SNVs=mtx_gene_sample_SNVs[i,])
  dfl <- melt(df,id.vars = c("sample_ID","MSI_status","total_mutations"))
  plot <- ggplot(dfl,aes(x=reorder(sample_ID,-total_mutations),y=value,fill=factor(MSI_status),alpha=variable)) +
  geom_bar(stat="identity",size=0.2) + ylab("Frequency of mutations") + xlab("Tumor samples") + 
  ggtitle(paste('Distribution of Indels and SNVs in ', gene_list[i], ' in each tumor sample', sep='')) +
  scale_fill_manual(values=c("red3","navy","grey50")) +
  scale_alpha_discrete(range=c(0.3,0.8)) + theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))
  print(plot)
}
```